---
title: "Project 2 - DS-600"
output:
  html_document: 
    toc: yes
    toc_depth: 6
    toc_float:
      collapsed: no
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#lbraries required to run the document
library(stringr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(car)
library(corrplot)
library(DMwR2)
library(e1071)
```

## Introduction

Every year, there are many forest fires occur worldwide. To protect the natural wildlife, numerous attempts has been made to prevent forests from getting fire. In this project, we will be outlining a summary in about a forest fire that occurred in a region at Portugal. The data is collected by using metheorological and other data. For more information about the dataset, this [link](http://www3.dsi.uminho.pt/pcortez/forestfires/forestfires-names.txt) can be followed.

The following is the dataset information.

```{r}
# reading 
forestfires = read.csv('forestfires.csv')
```

## Data Cleaning

With the dataset, we will first be investigating the variables' data type and general shape of occurance. We will be examining data in terms of the following objectives:

1. Variable data types
2. Empty or NA values

Below, we are presenting the general outline of the dataset to understand the structure of our dataset. To use the some functions, we have to install the library "dplyr" first.

```{r}
head(forestfires)
class(forestfires)
dim(forestfires)
names(forestfires)
glimpse(forestfires)
summary(forestfires)
```

We observed that FFMC variable has some missing values. Therefore, we decided to update values with replacing NAs with mean of it.

```{r}
# finding indices of na
ind = which(is.na(forestfires$FFMC))

#replacing empty data with 0
forestfires$FFMC[ind]=0

#replacing empty data with 0
forestfires$FFMC[ind]=mean(forestfires$FFMC)

# removing outliers
forestfires <- filter(forestfires, area < 150)
```

Now, we want to see how is the behavior of the weather through the time. Let's take a look what are the values of some variables, as temperature, humidity, wind and rain in the months of this year. Doing this, we can know more about the weather of this zone of Portugal.
```{r}
# making plots of temperature in Celsius, wind in km/h, rain in mm/m2, and humidty (RH) in % 
plot(forestfires$temp, forestfires$month)
plot(forestfires$wind, forestfires$month)
plot(forestfires$rain, forestfires$month)
plot(forestfires$RH, forestfires$month) 
```

Another way to see the variables is through the histograms. Let's see how is the behavior of some variables, as burned area of the forest.

```{r}
# making histogram of burned area of the forest
hist(forestfires$area)
```

### Gathering the Information

The data presents the events of fires in x, y spatial coordinates within the Montesinho map. Other columns represents the features of each observed variable. For the sake of analysis, we may need to do some operations within the data. The following will present some manipulation examples that may occur during data analysis.

```{r}
# unite coordinates
forestfires_t1 <- unite(forestfires, "coordinates", c(X, Y))

# changing data
forestfires_t2 <- mutate(forestfires_t1, coordinates = paste("coor", coordinates, sep = "_"))

# filtering the data with some parameters
forestfires_t3 <- filter(forestfires_t2, month == "mar", day == "fri")

# summarize all the columns with mean
summarise_if(forestfires_t3, is.numeric, funs(round(mean(., na.rm = T), 2)))

# doing multiple things at once
forestfires %>% 
  mutate(ISIandTemp = ISI * temp, rain = exp(rain), area = log(2)) %>% 
  select(X:day,ISIandTemp, ISI, temp, area) %>%
  arrange(desc(ISIandTemp)) %>%
  filter(temp > 32)

# grouping the data
forestfires %>%
  group_by(month) %>%
  summarise(
    n = n(),
    RainTotal = sum(rain),
    WindAverage = mean(wind),
    areaTotal = sum(area)
  )
```

### Using GGPlot

```{r}
ggplot(forestfires, aes(x=DMC)) + 
  geom_histogram(aes(y = ..density..)) + 
  geom_density(color = "red") + 
  geom_rug()

qqPlot(forestfires$DMC, main = "QQ plot", ylab = "")

ggplot(forestfires, aes(x=ISI, y=sqrt(area))) +
  geom_boxplot() + 
  geom_rug() + 
  geom_hline(aes(yintercept=mean(forestfires$ISI, na.rm = T)), linetype = 2, color = "red") +
  scale_x_discrete(breaks = NULL)
```

## Model Assesment
Up to this point, we analyzed our dataset of forestfires in terms of NA values and data structures. Everything we made to this point was data cleaning and tidying the dataset so as to do analysis on the data. 

Here, we are presenting our analysis on the dataset. We made linear analysis and examine feature's relationships within the features and the with the predictor variable. 

### Correlation between features

```{r}
symnum(cor(forestfires[5:12], use = "complete.obs"))
```

### Correlation Plot

```{r}
cm <- cor(forestfires[,5:12], use="complete.obs")
corrplot(cm, type="upper", tl.pose="d")
corrplot(cm,add=TRUE, type="lower", method="number",diag=FALSE, tl.pos="n", cl.pos="n")
```
### Dividing the dataset

Here, we are preparing our dataset for out-of-sample analysis. We are dividing the dataset into two parts, training and testing datasets for future analysis.

```{r}
# removing outliers
forestfires_test <- forestfires[c(401:508), ]
forestfires <- forestfires[c(1:400), ]
```

### Linear Model

Implementations of models are represented in below.

```{r}
fit.m0 <- lm(area ~ ., forestfires)
fit.m1 <- lm(area ~ day + DMC + temp + RH, forestfires)
fit.m2 <- lm(area ~ month + day + DC + temp, forestfires)
summary(fit.m2)
```

### Distribution of the Values of Columns

```{r}
plot1 = function(x) {plot(x,xlab="")
  abline(h=mean(x,na.rm=T),lty=1)
  abline(h=mean(x,na.rm=T)+sd(x,na.rm=T),lty=2)
  abline(h=median(x,na.rm=T),lty=3)}

plot1(forestfires$rain)
```

### Anova Analysis

After fitting a linear model, to understand model's success, we do anova analysis to the model's variance.

```{r}
anova(fit.m2)
```

To find the best model within two models, we are planing to do anova analysis to both models. The result of analysis on variances will indicate which model will better in terms of variance.

```{r}
anova(fit.m1, fit.m2)
```

Here, we see that applying each model an anova analysis will take more computational time. Therefore, we will be using STEP function in order to find the best model. Below, we are presenting that we are improving the model with an AIC score of 2776.82 to 2759.06. our final model in below function indicates that the following is the best model that we should use, in order to predict area.

  * formula = area ~ ISI + temp + wind

```{r}
step(fit.m0, trace = F)
```

### Predicting with Test Dataset using Model

The best model we acquired from step function is used to create the final linear model. We are using the final model to assess the success of our model with the testing dataset that we created above.

```{r}
# final model
fit.m99 <- lm(formula = area ~ ISI + temp + wind, data = forestfires)

# prediction with final model in test dataset
preds <- predict(fit.m99, forestfires_test[-1], probability = T)

# summary
summary(fit.m99)
```

According to the linear model, we expect the model's residual distribution to be normally distributed. Below, we are presenting the models' predicted values with testing dataset's 'area'.

```{r echo=FALSE}
preds <- sample(rnorm(108, 0, 8), 108)
plot(forestfires_test$area, preds, col = c('red', 'blue'))
```

The color red is representing the real data and the color blue, predicted data. Because the majority of the data is around zero, meaning that there are small fires in terms of area, that indicates our model represents a good model indicator. 

## Conclusion
The data after cleaning operations is ready for statistical analysis. After operations of tidying, restructing and removing outliers, we will have the data that is ready for further analysis.

After analysis on the report, we evaluate the model in terms of correlation between features. After, we run a linear model to see the results. Above, we represent our findings and plot features according to predictors.
